---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

```{r echo = FALSE}
library(countyfloods)
```


## Overview of the package

## Basic example

You can use the `get_county_cd` function to get a vector of all counties within a state: 

```{r}
get_county_cd(state = c("Georgia", "Alabama"))
```

You can use the `get_gages` function to pull all gages within a county or counties. For example, to get information on all gages for Miami-Dade county, you can run: 

```{r message = FALSE}
library(dplyr)
get_gages("12086", start_date = "1988-01-01", end_date = "2015-01-01") %>%
  slice(1:5)
```

You can use these two functions together within a pipe chain. For example, to get information on all the gages in Virginia in 2015, you can run: 

```{r message = FALSE}
va_gages <- get_county_cd("Virginia") %>%
   get_gages(start_date = "2015-01-01", end_date = "2015-12-31")
head(va_gages)
```

Once you have a list of gage numbers for which you would like to pull flow data, you can use the `get_flow_data` function to pull that data. For example, to pull data for all the stream gage flow data for 1999 for gages in Southhampton County, VA (FIPS code 51175), you can run: 

```{r}
southampton_gages <- get_gages("51175", start_date = "1999-01-01",
                                end_date = "1999-12-31")
southampton_data <- get_flow_data(southampton_gages, start_date = "1999-01-01",
                                  end_date = "1999-12-31")
```

The output from `get_flow_data` is a data frame with three columns ("site_no" = gage number, "date" = date
of observation, and "discharge" = mean daily flow in cubic feet per second): 

```{r}
head(southampton_data)
```

```{r fig.width = 6, fig.height = 3.5}
library(ggplot2)
ggplot(southampton_data, aes(x = date, y = discharge, color = site_no)) + 
  geom_line()
```

To determine when a flood occurred, we need a value to classify a "flood" for each stream gage.

One way to get this is with the `find_nws` function, which gets the National Weather Service flood value when available for a gage site. For example, for the Virginia gages: 

```{r}
va_nws <- find_nws(site_no = va_gages$site_no, type = "moderate")
head(va_nws)
dim(va_nws)
```

Another way is with the `find_q2` function which calculates the median annual flood for each gage using
a minimum of 20 years of annual peak flow data:

```{r}
va_q2 <- find_q2(site_no = va_gages$site_no)
head(va_q2)
dim(va_q2)
```

You can compare the results from these two methods for sites where you can get both values: 

```{r fig.width = 4, fig.height = 3.5, fig.align = "center"}
va_flood_stage <- va_nws %>%
  rename(flood_nws = flood_val) %>%
  inner_join(va_q2, by = "site_no") %>%
  rename(flood_q2 = flood_val)
ggplot(va_flood_stage, aes(x = flood_q2, y = flood_nws)) + 
  geom_point(alpha = 0.5) + 
  geom_abline(aes(intercept = 0, slope = 1), linetype = 3) + 
  geom_smooth(method = "lm", se = FALSE)
```

For the Virginia monitors, you can see that the flood values from NWS and the Q2 method are well-correlated, although NWS values tend to be consistently higher than Q2 values.

Once you have data on gages, flood values, and flow data, you can use the `flood_analysis` function to get flood summaries by site. For example, to get flood summary statistics by gage for the Virginia gages, you can run: 

```{r}
va_counties <- get_county_cd("Virginia")
va_flow_data <- get_flow_data(va_gages, start_date = "2015-01-01",
                              end_date = "2015-12-31")
va_floods <- flood_analysis(flow_data = va_flow_data, peaks = va_q2,
                            gages = va_gages, county_cd = va_counties,
                            threshold = "Q2")
head(va_floods, 3)
```

```{r fig.width = 6, fig.height = 3.5, fig.align = "center", message = FALSE}
ggplot(va_floods, aes(x = max_peak, fill = flood)) + 
  geom_histogram()
```

There is also a function that will allow you to get county-level aggregate statistics from this gage-level summary.
Any counties with no gages or for which the gages didn't have flow data or flood threshold values are also 
included: 

```{r}
va_county_stats <- county_aggregates(flood_stats = va_floods)
head(va_county_stats)
```

You can use the `run_flood` function to put all of this together, and pull all flood summaries by either gage (`output = "gage"`), county (`output = "county"`), or both (`output = "both"`) for either a set of counties or all counties in a state. It is more efficient to set (`output = "both"`) and then extract gage or county level data
from the list than running each seperately since pulling flow data is time consuming.

For example, to get all the flood statistics by gage for all gages with available data in Virginia, you can run: 

```{r message = FALSE}
va_floods <- run_flood(state = "Virginia", start_date = "2015-01-01",
                       end_date = "2015-12-31", threshold = "Q2",
                       output = "gage")
head(va_floods)
```

Similarly, to get county-level data for counties in Florida in 2004, you can run:

```{r message = FALSE}
fl_floods <- run_flood(state = "Florida", start_date = "2004-01-01",
                       end_date = "2004-12-31", threshold = "Q2",
                       output = "county")
head(fl_floods)
```

These output can be mapped using the `map_flood` function. If the data was collected by gage, this will show a point map with the flood level at each gage:

```{r fig.width = 8, fig.height = 3.5, fig.align = "center", message = FALSE}
map_flood(va_floods) 
```

For county-level data, this will create a choropleth indicating the percent of monitors in each county with flood magnitude above a user-specified flood category: 

```{r fig.height = 3, fig.width = 4.5, fig.align = "center", message = FALSE}
map_flood(fl_floods)
map_flood(fl_floods, category = "major")
```

The `long_term_flood` function is very similar to the `run_flood` function except it accepts a data frame as
input with three columns: `county_cd`, `start_date`, and `end_date`. This allows you to analyze floods across multiple date ranges and multiple counties. For example, if we wanted to examine April flooding in three counties
in northern Virginia we would create the following input data frame:

```{r message = FALSE}
county_cd <- c(rep("51013", 5), rep("51107", 5), rep("51059", 5))
start_date <- rep(c("2010-04-01", "2011-04-01", "2012-04-01", "2013-04-01", "2014-04-01"), 3)
end_date <- rep(c("2010-04-30", "2011-04-30", "2012-04-30", "2013-04-30", "2014-04-30"), 3)
input_df <- data.frame(county_cd = county_cd, start_date = start_date, end_date = end_date, stringsAsFactors = FALSE)
head(input_df)
```

It is important that these variables are character strings and not factors. The flood analysis can then be 
performed:

```{r message = FALSE}
#With default values
va_floods <- long_term_flood(input_df)
va_gage_output <- va_floods[[1]]
head(va_gage_output)
va_county_output <- va_floods[[2]]
head(va_county_output)
```

If you are interested in seeing when flooding occurred over a set time period, you can perform a time series
analysis using the `time_series_flood` function. This has similar inputs to the `run_flood` function. For
example, if you wanted to examine when flooding occurred in Virginia from 2010 to 2015:

```{r message = FALSE}
va_time_series <- time_series_flood(state = "Virginia", start_date = "2010-01-01",
                      end_date = "2015-12-31", threshold = "NWS",
                      flood_type = "flood")
va_gage_output <- va_time_series[[1]]
head(va_gage_output)
va_county_output <- va_time_series[[2]]
head(va_county_output)
```

The county-level output can be plotted using the `time_series_plot` function which shows bar charts of the
timing and magnitude of floods during the selected time period. You can select values for `start_date` and `end_date` to change the x-limits on the plots or the default is to show the full time period including any 
flood.

```{r fig.height = 6, fig.width = 8, fig.align = "center", message = FALSE}
time_series_plot(va_county_output)
```

## More detailed examples

Hurricane Floyd made landfall on Sept. 16, 1999, in North Carolina and caused extensive flooding, especially in eastern North Carolina. Here are maps for the month from Sept.15, 1999 to Oct. 15, 1999:

```{r fig.width = 7.5, fig.height = 2.75, fig.align = "center", message = FALSE}
nc_floods <- run_flood(state = "North Carolina", start_date = "1999-09-15",
                       end_date = "1999-10-15", threshold = "Q2",
                       output = "both")
nc_maps <- map_flood(nc_floods)
nc_maps[[1]]
nc_maps[[2]]
```

You can use the `map_tracks` function from `hurricaneexposure` (currently available in a development version on GitHub), using the flood maps as the base `plot_object` in this call. For example:

```{r fig.height = 3.5, fig.width = 6.25, fig.align = "center", message = FALSE}
library(hurricaneexposure)
map_tracks(storms = "Floyd-1999", plot_object = nc_maps[[2]])
```

```{r fig.height = 3.5, fig.width = 6.25, fig.align = "center", message = FALSE}
map_tracks(storms = c("Bonnie-2004", "Charley-2004", 
                      "Frances-2004", "Jeanne-2004"), 
           plot_object = map_flood(fl_floods, category = "major"),
           color = "aquamarine3")
```




