---
title: "Using the `countyfloods` package"
author: "Rod Lammers and Brooke Anderson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo = FALSE}
library(countyfloods)
```


## Overview of the package

## Basic example

You can use the `get_county_cd` function to get a vector of all counties within a state: 

```{r}
get_county_cd(state = c("Georgia", "Alabama"))
```

You can use the `get_gages` function to pull all gages within a county or counties. For example, to get information on all gages for Miami-Dade county, you can run: 

```{r}
get_gages("12086", start_date = "1988-01-01", end_date = "2015-01-01")
```

You can use these two functions together within a pipe chain. For example, to get information on all the gages in Virginia in 2015, you can run: 

```{r}
va_gages <- get_county_cd("Virginia") %>%
   get_gages(va_counties, start_date = "2015-01-01", end_date = "2015-12-31")
head(va_gages)
```

Once you have a list of gage numbers for which you would like to pull flow data, you can use the `get_flow_data` function to pull that data. For example, to pull data for all the stream gage flow data for 1999 for gages in Southhampton County, VA (FIPS code 51175), you can run: 

```{r}
southampton_gages <- get_gages("51175", start_date = "1999-01-01",
                                end_date = "1999-12-31")
southampton_data <- get_flow_data(southampton_gages$site_no, start_date = "1999-01-01",
                                  end_date = "1999-12-31")
```

The output from `get_flow_data` is a list with as many elements as there are gages with non-missing data: 

```{r}
class(southampton_data)
length(southampton_data)
```

Each element of this list gives streamflow data for one of the gages: 

```{r}
head(southampton_data[[1]])
```

```{r}
ggplot(southampton_data[[1]], aes(x = Date, y = Discharge)) + 
  geom_line()
```




## More detailed examples
