---
title: "Using the `countyfloods` package"
author: "Rod Lammers and Brooke Anderson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo = FALSE}
library(countyfloods)
```


## Overview of the package

## Basic example

You can use the `get_county_cd` function to get a vector of all counties within a state: 

```{r}
get_county_cd(state = c("Georgia", "Alabama"))
```

You can use the `get_gages` function to pull all gages within a county or counties. For example, to get information on all gages for Miami-Dade county, you can run: 

```{r}
get_gages("12086", start_date = "1988-01-01", end_date = "2015-01-01")
```

You can use these two functions together within a pipe chain. For example, to get information on all the gages in Virginia in 2015, you can run: 

```{r}
va_gages <- get_county_cd("Virginia") %>%
   get_gages(va_counties, start_date = "2015-01-01", end_date = "2015-12-31")
head(va_gages)
```



## More detailed examples
